\chapter{Conclusion}

\label{Conclusion}

This thesis proposes a solution for computer aided assistance in commonly used diagnostics in autoimmune diseases. The solution closely follows the procedure medical experts suggest. It covers the segmentation task,  the fluorescence intensity level classification and staining pattern classification. The strongest contribution of this thesis is the interpretable approach to prediction models that help medical experts. \\

The work on the segmentation part was motivated by two problems found in previous approaches. Due to the coloring procedure of fluorescence imaging, cells exhibit different properties which makes them hard to detect, especially in noisy images. By finding the background instead of cells directly, that problem was successfully solved with great improvement. Still, a lot of cells overlap.  By using the information about the background, morphological snakes were introduced. Although not always capable of separating overlapping cells, the method demonstrated very promising results. The problematic case were the patterns with bright organelles where those organelles were segmented. \\

The cells were then characterized with \textit{visual concepts} describing their appearance. Deep belief networks were used to map images, or pixel features to \textit{symbolic} ones. Deep approaches to machine learning demonstrated the great potential in symbolic feature recognition. Surprisingly, convolutional components seem not to improve the recognition, but the full potential of this approach is yet to be investigated. \\

The symbolic representations learned in the previous step are then used to mine rules that describe the patterns. Four commonly used rule mining algorithms were compared and proven to perform as accurately as state-of-the-art methods, but offer explanations about their decisions. I believe such an approach is needed when computers assist in making a diagnosis, based on images or other diagnostic tests.

\begin{flushleft}
	\large
	\vspace{15pt}
	\textbf{Future work}
\end{flushleft}

In order to make a fully functional system that supports this process, there are still problems that need to be solved. The segmentation task still suffers from certain problems, such as \textit{overexposed} organelles in a cell's body which end oversegmented. Introducing a model based segmentation method, where movement of the contour is restrained by model learned from data, might be a way to improve the segmentation of cells. \\

There are two actions that have not been mentioned in this thesis, but are of high interest to this task. Mitotic cells were taken as ground truth so far, but detecting and classifying them is a very important task for the diagnostic procedure. Together with mitotic cells, images taken by fluorescence imaging often have artifacts that don't carry any information. Both mitotic cells and artifacts are very rare objects in images which is the main difficulty of this task. Approaching this problem from an unbalanced classification or an outlier detection perspective could be a good start. \\

Symbolic feature learning achieved by deep belief network has proven to have a great potential. As deep learning demonstrates an enormous growth in research, experimenting with different deep approaches to improve symbolic representation learning leaves a lot of space for future work. Special emphasis should be put on methods that can deal with unbalanced classes which usually represent very significant features.